Increasing Productivity Amid Stable Nutrient Regimes in Rhode Island Lakes and Reservoirs

### 2019-05-08 manuscript/manuscript.Rmd
abstract done! closes #42

---
title: "Increasing Productivity Amid Stable Nutrient Regimes in Rhode Island Lakes and Reservoirs"
author:
- Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers,
  S. ^*3*^, Milstead, W. Bryan. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A.
  ^*2*^
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Ecology Division, Narragansett, RI 02882
- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*3*^ ORISE, Narragansett, RI 02882
- '^\*^  *corresponding author: hollister.jeff@epa.gov*'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: plos.csl
abstract: ""
---

@@

The anomaly approach is very useful and informed the approach we chose to follow.  However, using anomalies works best with a single measure (e.g temperature) or with multiple measurements that are on the same scale.  The water quality parameters that we explored have different scales and thus the anomaly alone is difficult to interpret across metrics.  For instance, temperature in Rhode Island lakes during the growing season will range from approximately 15 degrees Celsius to a high of 30 degrees Celsius, whereas phosphorus might range from near zero ug/l to ~900 ug/l.  To standardize these values we used the common approach of dividing each anomaly by the standard deviation for the reference period (e.g., [@jones1996calculating]).  The resultant value is commonly referred to as a z-score.  We use these z-scores to examine each water quality variable for a trend over the time period of 1993 to 2016.   Furthermore, since we are interested in water quality trends over time, we wanted to explore how each site was responding.  Thus, our z-scores were calculated over the reference period, 1993-2016, for each site.

### Summarizing z-scores

Details for our approach of calculating the yearly z-scores are as follows and are presented graphically in NEED FIGURE FOR THIS.  The steps below are repeated for each of the water quality parameters.

1. For each site, calculate the reference mean and reference standard deviation for all values.
2. Calculate the z-score for each measurement by subtracting the reference mean and dividing by the reference standard deviation.
3. Summarize the z-scores by calculating an average z-score for each site in each year.  This step prevent bias from pseudoreplication of multiple measurements of the same site in a given year ([@hurlbert1984pseudoreplication]).  The same site averages across years are assumed to be independent.
4. Summarize the yearly z-scores by calculating the average, standard deviation, and sample size of the per site z-scores.

>>>>

---
title: "Increasing Productivity Amid Stable Nutrient Regimes in Rhode Island Lakes and Reservoirs"
author:
- Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers,
  S. ^*3*^, Milstead, W. Bryan. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A.
  ^*2*^
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Ecology Division, Narragansett, RI 02882
- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*3*^ ORISE, Narragansett, RI 02882
- '^\*^  *corresponding author: hollister.jeff@epa.gov*'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: plos.csl
abstract: 'Addressing anthropogenic impacts on aquatic ecosystems has long been the focus of lake management. Controlling phosphorus and nitrogen can mitigate the impacts of eutrophication, but to determine its effectiveness requires long-term datasets.  A recent analysis of the LAke multi-scaled GeOSpatial and temporal database (LAGOSNE), found stable water quality in the Northeast and Midwestern regions of the United States, however, trends at smaller scales may be masked.  We address this by analyzing the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset.  URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years. These data, included in LAGOSNE, allow for comparison of water quality trends at regional and state extents. We assess for trends with z-scores (i.e. scaled anomalies) calculated on a per-station basis and examine yearly averages.  Temperature and chlorophyll *a* are increasing.  Total nitrogen shows a weak increasing trend driven by low years in the early 1990s.  Total phosphorus and the nitrogen:phosphorus ratio (N:P) were stable.  Applying the site-specific z-score approach to LAGOSNE found similar trends to prior studies with chlorophyll *a*, total nitrogen, total phosphorus, and N:P all stable over time.  In short, productivity in Rhode Island lakes and reservoirs is increasing, in spite of stable nutrient regimes. Although not causal, this analysis suggests an association between lake temperature and productivity.  Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific z-scores for analyzing for long-term trends. '
---

@@

The anomaly approach is very useful and informed the approach we chose to follow.  However, using anomalies works best with a single measure (e.g temperature) or with multiple measurements that are on the same scale.  The water quality parameters that we explored have different scales and thus the anomaly alone is difficult to interpret across metrics.  For instance, temperature in Rhode Island lakes during the growing season will range from approximately 15 degrees Celsius to a high of 30 degrees Celsius, whereas phosphorus might range from near zero ug/l to ~900 ug/l.  To standardize these values we used the common approach of dividing each anomaly by the standard deviation for the reference period (e.g., [@jones1996calculating]).  The resultant value is commonly referred to as a z-score.  We use these z-scores to examine each water quality variable for a trend over the time period of 1993 to 2016.   Furthermore, since we are interested in water quality trends over time, we wanted to explore how each site was responding.  Thus, our z-scores were calculated over the reference period, 1993-2016, for each site.  We refer to this approach as the site-specific z-scores

### Summarizing site-specific z-scores

Details for our approach of calculating the site-specific z-scores and the yearly averages are as follows and are presented graphically in `r fig_num(name = "z_schematic", caption = "Example calculation of the site-specific z-scores and yearly mean z-scores.", display = "cite")`.  Additionally, an example R script, `schematic.R` and example dataset, `schematic.csv` to recreate and demonstrate these calculations is available from <https://github.com/jhollist/ri_wq_trends>.  The general steps, listed below, are repeated for each of the water quality parameters.

WRONG ORDER JEFF!
1. For each site, calculate the reference mean and reference standard deviation for all values.
2. Calculate the z-score for each measurement by subtracting the reference mean and dividing by the reference standard deviation.
3. Summarize the z-scores by calculating an average z-score for each site in each year.  This step prevent bias from pseudoreplication of multiple measurements of the same site in a given year ([@hurlbert1984pseudoreplication]).  The same site averages across years are assumed to be independent.
4. Summarize the yearly z-scores by calculating the average, standard deviation, and sample size of the per site z-scores.
